---
title: "The impact of storms and severe weather events on economy and population health in USA"
output: 
  html_document:
    keep_md: true
---

```{r, echo = FALSE}
library(ggplot2)
```

## Synopsis
TBD: Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

## Data Processing
TBD: There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.


### Reading data

```{r reading, cache = TRUE}
# read.csv can open csv files from bz2 directly
storms <- read.csv("repdata-data-StormData.csv.bz2")
```

### Cleaning data
Quick look at the data shows that it needs tidying. Also it is clear that we need only a number of columns from this dataset to answer the questions of this analysis, so we'll perform some transformations described below.
```{r}
str(storms)
```
First of all, we do not need the entire dataset for several reasons:  
- According to [http://www.ncdc.noaa.gov/stormevents/details.jsp](Storm Events Database description) there are three distinct periods when data reporting changed significantly.  
For the analysis only third period will be selected which started in January 1996 and contains data for 48 event types recorded as defined in NWS Directive 10-1605. If we decided to use the complete dataset, results of the analysis may be significantly biased due to major differences in number and types of events reported during mentioned periods.  
- We need only events which led to **fatalities** or **injuries** to answer first question.  
- We need only events which caused **property damage** or **crop damage**.  

In order to get the subset of the data based on date, we need to convert BGN_DATE to Date format and then select events which began later than January, 1996.

```{r}
storms <- transform(storms, BGN_DATE = as.Date(as.character(BGN_DATE), "%m/%d/%Y"))
storms1996 <- subset(storms, BGN_DATE >= as.Date("1996-01-01"))
```

Select events with either fatalities or injuries or property damage or crop damage.
```{r}
selected <- subset(storms1996, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)
```

It is necessary to calculate damage value with the same base. In order to do this we need to multiply PROPDMG and CROPDMG based on PROPDMGEXP and CROPDMGEXP modifiers.
Look at damage values modifiers:
```{r}
table(selected$PROPDMGEXP)
table(selected$CROPDMGEXP)
```
We can see that only "K", "M", and "B" are used and they stand for one thousand, one million and one billion.

Adjust damage values using PROPDMGEXP and CROPDMGEXP modifiers.
```{r}
selected$PROPDMG[selected$PROPDMGEXP == "K"] <- 
    selected$PROPDMG[selected$PROPDMGEXP == "K"] * 1000
selected$PROPDMG[selected$PROPDMGEXP == "M"] <- 
    selected$PROPDMG[selected$PROPDMGEXP == "M"] * 1000000
selected$PROPDMG[selected$PROPDMGEXP == "M"] <- 
    selected$PROPDMG[selected$PROPDMGEXP == "M"] * 1000000000

selected$CROPDMG[selected$CROPDMGEXP == "K"] <- 
    selected$CROPDMG[selected$CROPDMGEXP == "K"] * 1000
selected$CROPDMG[selected$CROPDMGEXP == "M"] <- 
    selected$CROPDMG[selected$CROPDMGEXP == "M"] * 1000000
selected$CROPDMG[selected$CROPDMGEXP == "M"] <- 
    selected$CROPDMG[selected$CROPDMGEXP == "M"] * 1000000000
```

Next step is to clean levels of EVTYPE. Using subsetting we reduced the number of distinct EVTYPES and we need to adjust levels.

```{r}
# adjust levels and make them uppercase
selected$EVTYPE <- factor(toupper(as.character(selected$EVTYPE)))
dirtyevents <- as.character(levels(selected$EVTYPE))
dirtyevents
```

Even after subsetting we can see a huge number of weather events compared to expected 48 events listed in [http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf](NWS Directive 10-1605)  
EVTYPE needs to be cleaned.

Many event names in the dataset have the following defects:  
- short versions of the official event names  
- names with addidional suffexes  
- names with spaces in inappropriate places  
- combined events  
- names made up with no reference to official event names  

```{r}
# Events from the official document NWS Directive 10-1605
events <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")

# make upper-case version of event names
eventsUC <- toupper(events)
```

Event names as described in the official document that are found in the dataset:
```{r}
intersect(as.character(levels(selected$EVTYPE)), eventsUC)
```

Event names which are different from the official name or have additional suffixes:
```{r}
setdiff(as.character(levels(selected$EVTYPE)), eventsUC)
```

Modifying event names to match the ones from the official document:
```{r}
# This is to map all names needed for quantile 0.9 of totals
levels(selected$EVTYPE)[grep("TSTM WIND", 
                             as.character(levels(selected$EVTYPE)))] <- "THUNDERSTORM WIND"
levels(selected$EVTYPE)[grep("RIP CURRENTS", 
                             as.character(levels(selected$EVTYPE)))] <- "RIP CURRENT"
levels(selected$EVTYPE)[grep("EXTREME COLD", 
                             as.character(levels(selected$EVTYPE)))] <- "EXTREME COLD/WIND CHILL"
levels(selected$EVTYPE)[grep("^FOG", 
                             as.character(levels(selected$EVTYPE)))] <- "DENSE FOG"
levels(selected$EVTYPE)[grep("HURRICANE/TYPHOON", 
                             as.character(levels(selected$EVTYPE)))] <- "HURRICANE (TYPHOON)"
levels(selected$EVTYPE)[grep("WILD/FOREST FIRE", 
                             as.character(levels(selected$EVTYPE)))] <- "WILDFIRE"
levels(selected$EVTYPE)[grep("HURRICANE$", 
                             as.character(levels(selected$EVTYPE)))] <- "HURRICANE (TYPHOON)"
levels(selected$EVTYPE)[grep("STORM SURGE", 
                             as.character(levels(selected$EVTYPE)))] <- "STORM SURGE/TIDE"
levels(selected$EVTYPE)[grep("TYPHOON", 
                             as.character(levels(selected$EVTYPE)))] <- "HURRICANE (TYPHOON)"
levels(selected$EVTYPE)[grep("FREEZE", 
                             as.character(levels(selected$EVTYPE)))] <- "FROST/FREEZE"
```

Make new data structures to address both questions of the analysis:
```{r}
fatalities <- tapply(selected$FATALITIES, selected$EVTYPE, sum)
injuries <- tapply(selected$INJURIES, selected$EVTYPE, sum)
propdamage <- tapply(selected$PROPDMG, selected$EVTYPE, sum)
cropdamage <- tapply(selected$CROPDMG, selected$EVTYPE, sum)
```

Since we're interested in the most harmful event types, we select only top harmful events.

```{r}
fatalities <- fatalities[fatalities > quantile(fatalities, 0.9)]
injuries <- injuries[injuries > quantile(injuries, 0.9)]
propdamage <- propdamage[propdamage > quantile(propdamage, 0.9)]
cropdamage <- cropdamage[cropdamage > quantile(cropdamage, 0.9)]
```


```{r}
# create data.frame with the results for convinient plotting
fatalities.df <- data.frame(casualties = as.vector(fatalities), 
                       event = as.factor(names(fatalities)), 
                       harm = "fatality")

injuries.df <- data.frame(casualties = as.vector(injuries), 
                         event = as.factor(names(injuries)), 
                         harm = "injury")

harm.df <- rbind(fatalities.df, injuries.df)

```

## Results
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r health_results}
qplot(event, 
      casualties , 
      data=harm.df, 
      geom = "bar", 
      stat = "identity", 
      fill = harm)
```  

2. Across the United States, which types of events have the greatest economic consequences?

```{r economy_results}
```
